Clustering

### 2024-05-25 data/03-clustering-and-profiling.qmd
Completed initial experiments developing accessibility profiles and demographic profiles of parcels.

Lets begin by loading the packages used for this:
```{r message = FALSE}
library(AppliedPredictiveModeling) # Functions and Data Sets for 'Applied Predictive Modeling'
library(caret) # Classification and Regression Training
library(ggparty)
library(ggpubr)
library(party)
library(plotly) # Create Interactive Web Graphics via 'plotly.js'
library(rpart) # Recursive Partitioning and Regression Trees
library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
library(skimr)
#library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
library(tidyverse) # Easily Install and Load the 'Tidyverse'

# library(formattable) # Create 'Formattable' Data Structures
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
# library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
# library(plotly) # Create Interactive Web Graphics via 'plotly.js'
# library(sf) # Simple Features for R
# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
# library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
# library(tidyverse) # Easily Install and Load the 'Tidyverse'
# library(tmap) # Thematic Maps
```

@@

Given these normalized attributes, a SOM can be trained using the function `trainSOM` from `SOMbrero`. Only one argument is needed, the input data `x.data`. Other arguments can be assigned by default. In what follows the dimension of the map is given as 5-by-5, and I request that 100 intermediate steps be saved. 
```{r}
set.seed(593)
# run the SOM algorithm 
nrows <- 5
ncols <- 5
access.som <- trainSOM(x.data = access_15_wide |>
                       # Remove the non-normalized accessibility scores
                       select(-c(from_id, starts_with("total_"))),
                     dimension=c(nrows, ncols),
                     nb.save = 100,
                     radius.type = "letremy",
                     verbose = TRUE)
```

@@

Frequency of parcels in the super clusters (look at the data of the first node):
```{r}
freq_sc <- ggplot(classmod.sompar[1]$data) +
  geom_bar(aes(x = "", 
               fill = scluster)) +
  scale_fill_discrete(name = "super cluster") +
  theme_void()

@@

Extract the legend:
```{r}
leg <- ggpubr::get_legend(freq_sc,
                          position = "bottom")

leg <- as_ggplot(leg)

leg
```

Make a prettier plot:
```{r}
party_plot <- ggparty(classmod.sompar) +
  geom_edge() +
  geom_edge_label(size = 1.5) +
  geom_node_splitvar(size = 2) +
  # pass list to gglist containing all ggplot components we want to plot for each
  # (default: terminal) node
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = scluster),
                                        position = position_fill()),
                               xlab(""),
                               ylab(""),
                               theme_void(),
                               theme(axis.text = element_blank())))
```

@@

>>>>

Lets begin by loading the packages used for this:
```{r message = FALSE}
library(AppliedPredictiveModeling) # Functions and Data Sets for 'Applied Predictive Modeling'
library(caret) # Classification and Regression Training
library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2'
library(ggparty) # 'ggplot' Visualizations for the 'partykit' Package
library(party) # A Laboratory for Recursive Partytioning
library(plotly) # Create Interactive Web Graphics via 'plotly.js'
library(rpart) # Recursive Partitioning and Regression Trees
library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
library(skimr) # Compact and Flexible Summaries of Data
library(sf) # Simple Features for R
library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
library(tidyverse) # Easily Install and Load the 'Tidyverse'
```

@@

Given these normalized attributes, a SOM can be trained using the function `trainSOM` from `SOMbrero`. Only one argument is needed, the input data `x.data`. Other arguments can be assigned by default. In what follows the dimension of the map is given as 5-by-5, and I request that 100 intermediate steps be saved. 
```{r}
set.seed(5935)
# run the SOM algorithm 
nrows <- 5
ncols <- 5
access.som <- trainSOM(x.data = access_15_wide |>
                       # Remove the non-normalized accessibility scores
                       select(-c(from_id, starts_with("total_"))),
                     dimension=c(nrows, ncols),
                     nb.save = 100,
                     radius.type = "letremy",
                     verbose = TRUE)
```

@@

Frequency of parcels in the super clusters:
```{r}
freq_sc <- ggplot(access_15_wide |> 
         mutate(scluster = factor(scluster))) +
  geom_bar(aes(x = "", 
               group = scluster, 
               fill = scluster)) +
  scale_fill_discrete(name = "super cluster") #+
  #theme_void() +
  #guides(fill = guide_legend(nrow = 1)) +
  #theme(legend.position = "bottom")

@@

Extract the legend:
```{r}
legd <- cowplot::get_legend(freq_sc)

legd
```

Make a prettier plot:
```{r}
party_plot <- ggparty(classmod.sompar) +
  geom_edge() +
  geom_edge_label(size = 1.8) +
  geom_node_label(aes(label = splitvar),
                  id = "inner", 
                  size = 2, nudge_y = -0.022) +
  # pass list to gglist containing all ggplot components we want to plot for each
  # (default: terminal) node
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = scluster),
                                        position = position_fill()),
                               xlab(""),
                               ylab(""),
                               theme_void(),
                               theme(axis.text = element_blank())))
```

@@

```{r}
library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2' # Streamlined Plot Theme and Plot Annotations for 'ggplot2'

plot_grid(party_plot, legd, nrow = 1, rel_widths = c(1, 0.2))

```

Cluster 6 is high in suckitude: low in diversity, with 4.6 Dependent Centric destinations on average, and typically no other destinations: zero Errand centric destinations, and only about 3.5% of parcels have access to Grocery-centric destinations.
```{r}
access_15_wide |>
  select(scluster, starts_with("total_"), div) |>
  rename_with(.fn = ~str_replace(.x, "total_", "")) |>
  filter(scluster == 6) |>
  skim()
```

## Demographic profile of clusters

```{r}
load("HAM_census_21.rda")
```

```{r}
access_15_wide <- access_15_wide |>
  left_join(access_15 |>
              ungroup() |>
              distinct(from_id, 
                       .keep_all = TRUE) |>
              select(from_id, 
                     GeoUID),
            by = "from_id") |>
  left_join(HAM_census_21 |>
              st_drop_geometry(),
            by = "GeoUID")
```

Select data for classification:
```{r}
df <- access_15_wide |>
  transmute(scluster = factor(scluster),
            LICO = `v_CA21_1085: Prevalence of low income based on the Low-income cut-offs, after tax (LICO-AT) (%)`)
```

Train a decision tree using the data:
```{r}
classdemo.som <- rpart(scluster ~ ., 
                      data = df)
```

Visualize the decision tree:
```{r}
rpart.plot(classdemo.som, cex = 0.5, type = 0, varlen = 8)
```

convert the rpart object to a party object:
```{r}
classdemo.sompar <- as.party(classdemo.som)
```

Make a prettier plot:
```{r}
party_plot <- ggparty(classdemo.sompar) +
  geom_edge() +
  geom_edge_label(size = 1.8) +
  geom_node_label(aes(label = splitvar),
                  id = "inner", 
                  size = 2, nudge_y = -0.022) +
  # pass list to gglist containing all ggplot components we want to plot for each
  # (default: terminal) node
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = scluster),
                                        position = position_fill()),
                               xlab(""),
                               ylab(""),
                               theme_void(),
                               theme(axis.text = element_blank())))
```

```{r}
library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2' # Streamlined Plot Theme and Plot Annotations for 'ggplot2'

plot_grid(party_plot, legd, nrow = 1, rel_widths = c(1, 0.2))

```

Notice that super cluster 3 () makes most of the node with the highest prevalence of poverty (LICO > 10.9%). What is the accessibility profile of this cluster?

It is high diversity, with multiple Grocery with multiple Grocery- and Dependant-centric opportunities, and at least some Errand-centric opportunities:
```{r}
access_15_wide |>
  select(scluster, starts_with("total_"), div) |>
  rename_with(.fn = ~str_replace(.x, "total_", "")) |>
  filter(scluster == 3) |>
  skim()
```

It is clear that LICO alone is not sufficient to sharply profile the clusters. Ask Anastasia to choose other variables.
